---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# coxln

<!-- badges: start -->
<!-- badges: end -->

The goal of coxln is to simulate and model point pattern data on linear networks using various classes of Cox point processes, specifically:

- Log Gaussian Cox processes
- Interrupted Cox processes
- Permanental Cox point processes

The package contains different algorithms for simulating these Cox processes, as well as the underlying Gaussian processes. Furthermore it contains algorithms for calculating summary statistics (F, G, J and pcf) using either geodesic or resistance metric, which is useful for parameter estimation and model checking.

## Installation

You can install the development version of coxln from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("gulddahl/coxln",dependencies=TRUE,build_vignettes=TRUE)
```

## Examples

The first example shows how to simulate a log Gaussian Cox process with exponential covariance function and geodesic metric on a linear network called simplenet (from the spatstat package) using a grid on the network.

```{r fig.width=5, fig.height=4}
library(coxln)
pos = makepos(spatstat.data::simplenet,50)
covfunc = covfunctypes("expcov",1)
X = simCPLNDisc(pos,covfunc,sigma=1,transform="lgcp",rho=20)
plot(attr(X,"Lambda"),style="width",col="grey",main="Simulation of LGCP and underlying GRF")
plot(X,add=TRUE)
```

The second example shows how to estimate the F-function from the data set spiders (also from the spatstat package) using the resistance metric:

```{r fig.width=5, fig.height=4}
library(coxln)
X = spatstat.data::spiders
r = seq(0,250,length.out=50)
ppul = 0.02
F = linearF(X,r,ppul,metric="R")
plot(F)
```

The final example shows how to estimate the parameters of a log Gaussian Cox process with an exponential covariance function and the resistance metric on the spiders data using minimum contrast estimation based on the pair correlation function:

```{r message=FALSE, fig.width=5, fig.height=4}
library(coxln); library(spatstat)
r = seq(0,500,length.out=50)
gest = linearpcfR(spatstat.data::spiders,r)
g = paircorfunc("expcov","lgcp")
MCE = mincontrast(gest,g,c(0.1,0.1),ctrl=list(rmin=100,rmax=300))
param = c(sigma=MCE$par[1],s=MCE$par[2])
plot(gest,legend=FALSE,xlim=c(100,300),ylim=c(0.5,1.5))
curve(g(param,x),col="green",add=TRUE)
```

The estimated parameters are now stored in param, which in this case is given by (sigma,s) = (`r round(param,4)`), i.e. the standard deviation of the underlying Gaussian random field and the scaling parameter of the exponential covariance function.

# Vignette

An example of how to do a statistical analysis of a dataset using the functions from the coxln package is provided in the vignette for this package:

```{r eval=FALSE}
vignette("coxln")
```
