---
title: "coxln: Fitting a log Gaussian Cox process to a point pattern on a linear network"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coxln: Fitting a log Gaussian Cox process to a point pattern on a linear network}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of this vignette is to illustrate how to use the package for fitting a log Gaussian Cox process model to a point pattern dataset defined on a linear network, as well as giving an overview of alternative models and algorithms in the package.

# Loading packages

We start by loading the coxln package, and since we will be using much of spatstat's functionality and a couple of datasets we also load this package. Furthermore we will be using global envelope tests for model checking, so we will also need the GET package.

```{r message=FALSE}
library(coxln)
library(spatstat)
library(GET)
```

# Loading and viewing data

We will use the chicago dataset from the spatstat package to illustrate the methods. This contains positions of street crimes on a part of the Chicago street network. The dataset is marked with the type of crime, but for this analysis we will ignore the marks.

```{r fig.width=4,fig.height=4,fig.align='center'}
X = unmark(chicago)
L = as.linnet(X)
plot(X,pch=20,cols="red")
```

# Preliminary analysis

Before we define a model, we view the pair correlation function for a range of distances. In principle we can choose between the geodesic (or shortest path) metric and the resistance metric, but only the resistance metric leads to well-defined models later on, so we choose this metric.

```{r cache=TRUE, fig.width=4,fig.height=4,fig.align='center'}
r = seq(0,100,length.out=50)
gest = linearpcfR(X,r)
plot(gest,main="Pair correlation function")
```

Except for very small values of r, the pair correlation function is above 1, suggesting that the data is clustered, meaning that a Cox process could well be an appropriate model for this data. 

# Defining model

We will use a log Gaussian Cox process with an exponential covariance function 
\[
c(r) = \exp(-sr)
\]
with a scale parameter $s>0$ for the underlying Gaussian random field. This simply boils down to defining the theoretical pair correlation function as follows:

```{r}
g = paircorfunc("expcov","lgcp")
```

The resulting output g is a function taking a vector of the parameters of the pair correlation function as its first input, and the distance r as its second input. The parameters of the pair correlation function are given by $(\sigma,s)$, which is a subset of the parameters of the used log Gaussian Cox process model:

- $\sigma>0$ is the standard deviation of the underlying Gaussian random field.
- $s>0$ is the scale parameter used in the exponential covariance function.
- $\rho>0$ is the intensity of the log Gaussian Cox process.

The fact that $\rho$ is not included in the pair correlation function is dealt with later.

# Estimating parameters

For parameter estimation we use minimum contrast estimation based on the previously estimated pair correlation function. Since the covariance function of the underlying Gaussian random field has to be completely monotone for it to be well-defined on arbitrary linear networks, the theoretical pair correlation function is required to be decreasing (among other things), so we can never hope to catch the behavior of the estimated pair correlation function at ranges $r=[0;20]$, and thus we focus on the interval $r=[20;100]$ in the minimum contrast estimation procedure.

```{r}
MCE = mincontrast(gest,g,startpar=c(0.1,0.1),ctrl=list(rmin=20,rmax=100))
parest = MCE$par  # (sigma,s)
```

From this we obtain the parameter estimates $(\hat\sigma,\hat s) =$ (`r round(parest,4)`).
As a sanity check we can immediately compare the non-parametric estimate of the pair correlation function with the one corresponding the fitted model. 

```{r fig.width=4,fig.height=4,fig.align='center'}
plot(gest,legend=FALSE,xlim=c(20,100),ylim=c(0,3),main="Pair correlation function")
curve(g(parest,x),col="green",add=TRUE)
```

Since the pair correlation function does not depend on $\rho$, this parameter has to be estimated separately. Here we simply estimate it by the usual estimate of the intensity, i.e. the number of points divided by the total length of the network.

```{r}
rho = npoints(X)/volume(L)
```

This gives the estimate $\hat\rho=$ `r round(rho,5)`.

# Simulating the fitted model

The fitted model can now be simulated and compared to the data as a crude model check. To simulate this we discretize the linear network with approximately 0.02 grid points per unit length of network.

```{r fig.width=4,fig.height=4,fig.align='center'}
ppul = 0.02
pos = makepos(L,ppul)
plot(pos,pch=20,cols="red",cex=0.5)
```

This leaves us with a total of `r npoints(pos)` points in the grid. We can now make a simulation on this grid, where we first have to define the covariance function of the underlying grid (which obviously have to match the choice of pair correlation function previously chosen). We plot both the simulated point pattern and the underlying Gaussian random field.

```{r fig.width=5,fig.height=5,fig.align='center'}
covfunc = covfunctypes(par=parest[2],type="expcov")
sim = simCPLNDisc(pos,covfunc,sigma=parest[1],metric="R",transform="lgcp",rho=rho)
plot(attr(sim,"Lambda"),style="width",adjust=2,main="")
plot(as.ppp(sim),pch=20, cols="red",add=TRUE)
```

# Checking the fit of the model

For model checking we will compare summary statistics estimated from a number of simulations with the same summary statistic estimated from the data. We do this in the framework og the global envelope test. We have already used the pair correlation function for estimation, so we will use the F, G and J-functions for model checking. 
First we generate a number of simulations from the estimated model:
```{r eval=FALSE}
simlist = list()
nsim = 999
for (i in 1:nsim) {
  print(i)
  sim = simCPLNDisc(pos,covfunc,sigma=parest[1],metric="R",transform="lgcp",rho=rho,
                    savelambda=FALSE)
  simlist = c(simlist, list(sim))
}
```

Next we use the envelope function for calculating envelopes for use in the global envelope test:

```{r eval=FALSE}
EF = envelope(X,linearF,nsim=nsim,simulate=simlist,savefuns=TRUE,r=r,ppul=ppul,metric="R")
EG = envelope(X,linearG,nsim=nsim,simulate=simlist,savefuns=TRUE,r=r,metric="R")
EJ = envelope(X,linearJ,nsim=nsim,simulate=simlist,savefuns=TRUE,r=r,ppul=ppul,metric="R")
```

Finally we concatenate the F, G and J functions, and make a global envelope test based on the concatenated function:

```{r eval=FALSE}
EFGJglobal = global_envelope_test(list(EF,EG,EJ))
plot(EFGJglobal,scales="free")
```

```{r echo=FALSE, out.width=600, out.height=620, fig.align='center'}
knitr::include_graphics("chicago-FGJglobal-lgcp.pdf")
```

A warning is in order: Making 999 simulations and calculating envelopes for all three summary statistics is quite time-consuming and takes up a lot memory.

# Other point process models

The package also implements two other Cox processes: Interrupted Cox processes (ICP) and Permanental Cox point processes (PCPP), which both can be used with minor changes in the above example.

## ICP

To get an ICP, the "lgcp" should be exchanged with "icp" in the definition of the pair correlation function. The pair correlation function now depends on an additional parameter $h$, which is the number of Gaussian random fields used in the definition of an interrupted Cox process, i.e. now the parameter vector used in the pair correlation function is given by $(\sigma,h,s)$. 

The inclusion of the $h$ parameter changes the estimation procedure a bit, since it is integer valued (the mincontrast function assumes continuous parameters), and it has also been observed that it is difficult to distinguish between the effects of changing $h$ and $\sigma$. One option is to fix $h$ at a chosen value, for example $h=1$:
```{r}
g = paircorfunc("expcov","icp")
gh = function(par,r,...) g(c(par[1],h=1,par[2]),r) 
```
The estimation then proceeds as before using gh instead of g. The simulation algorithm simCPLNDisc accepts h as input when simulating an ICP.

## PCPP 

Here "lgcp" should be exchanged with "pcpp". In this model the parameter $h$ is also included, but $\sigma$ is not. Here it may be relevant to use several fixed values of $h$ and compare the fitted models to see which one fits the data best. Otherwise everything proceeds as before.

# Other covariance functions

Other covariance functions than the exponential can also be used. Currently three cases are implemented, all of which are defined by the corresponding Bernstein distribution.

## Covariance function with gamma Bernstein distribution

Here we use type="gamma" in the definition of the covariance function, i.e. we write the following to get a pair correlation function for a LGCP with gamma Bernstein distribution:

```{r}
g = paircorfunc("gamma","lgcp")
```

The covariance function is now given by 
\[
c(r) = (1+r/\phi)^{-\tau}
\]
where parameters are now given by a shape parameter tau and a rate parameter phi. 
The parameter vector that g depends on is always given by a vector containing first the parameters from the chosen Cox process and second the chosen covariance function. For example for an LGCP model with type="gamma", the parameter vector of the pair correlation function is given by $(\sigma,\tau,\phi)$, or for an similar ICP model it is given by $(\sigma,h,\tau,\phi)$. When estimating the parameters by using the mincontrast function, the startpar input should always be a vector of the same length as the parameter vector

Note that all the models also have the intensity parameter $\rho$, which never appears in the pair correlation function. Therefore this never enters the minimum contrast estimation procedure, but is estimated as in the previous analysis.

## Covariance function with inverse gamma Bernstein distribution

Here we use type="invgamma" to get the covariance function given by
\[
c(r) = 2(r \phi)^{\tau/2}K_\tau(2 \sqrt{r \phi}/(\Gamma(\tau))
\]
with shape parameter tau and scale parameter phi.

## Covariance function with generalized inverse Gaussian Bernstein distribution

Here we use type="gig":
\[
c(r) = (1+2r/\psi)^{-\lambda/2}\frac{K_\lambda(\sqrt{(2r+\psi)\chi})}{K_\lambda(\sqrt{\psi \chi})}
\] with parameter vector $(\chi,\psi,\lambda)$.

## A warning about non-exponential covariance functions

Non-exponential covariance functions work fine in the simulation algorithms, but the minimum contrast estimation does not work well. For all three choices of non-exponential covariance functions, the resulting parameter estimates usually gives something very close to an exponential covariance function, even when estimation is applied to simulated data using a clearly non-exponential covariance function. This is currently an unsolved problem. 

# Tree-shaped linear networks

For tree-shaped algorithms all the methods above work fine, but there is a much faster algorithm for simulation in the case of an exponential covariance function, which takes advantage of the fact the resulting process is Markov. For example, making a simulation of an LGCP on the network from the dendrite dataset in the spatstat package can be achieved in the following way:

```{r fig.width=5,fig.height=5,fig.align='center'}
L = as.linnet(dendrite)
pos = makepos(L,0.5,duplicate=TRUE)
sim = simCPExpLNRoot(pos,beta=0.01,sigma=1,transform="lgcp",rho=0.05)
plot(attr(sim,"Lambda"),style="width",adjust=2,main="")
plot(as.ppp(sim),pch=20, cols="red",add=TRUE)
```

Note that for technical reasons makepos needs the argument duplicates=TRUE in order for this algorithm to work.

For non-exponential covariance functions there is also another algorithm for trees, but this is only approximate, and whether or not it is faster than the algorithm used in the analysis above depends on the number of simulations used in the approximation and the number of points in the grid (for a sufficiently high number of grid points, the below algorithm becomes the fastest). For example, the algorithm in the case of an LGCP with a covariance function with a gamma Bernstein distribution is given by the following:

```{r fig.width=5,fig.height=5,fig.align='center'}
L = as.linnet(dendrite)
pos = makepos(L,0.5,duplicate=TRUE)
simalgo = simalgotypes(param=c(5,5),type="gamma",nsim=50)
sim = simCPLNRoot(pos,simalgo,sigma=1,transform="lgcp",rho=0.05)
plot(attr(sim,"Lambda"),style="width",adjust=2,main="")
plot(as.ppp(sim),pch=20, cols="red",add=TRUE)
```
